Iterative Monomorphisation

### 2024-02-28 iterative_monomorphisation.tex
clause monomorphisation step

\usepackage{soul}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
\usepackage%[linesnumbered]
  {algorithm2e}
\usepackage{amsmath}

@@

\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{algoColorKeyword}{named}{grey}
\makeatletter
% Block with a vertical line
\renewcommand{\algocf@Vsline}[1]{%
   \strut\par\nointerlineskip%
   \algocf@push{\skiprule}%
   \hbox{%
      {\color{algoColorKeyword}\vrule}
      \vtop{\algocf@push{\skiptext}%
      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
   }%
   \algocf@pop{\skiprule}%
}

@@

\begin{quote}
\begin{algorithm}[H]
   \caption{Substitution Generation}
%   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,formula \(\varphi\), monomorphic type argument map \(M\), non-monomorphic type argument map \(P\)}
   \Output{\,set of substitutions}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \Continue \(\leftarrow\) \True\;
         \(i \leftarrow 0\)\;
         \While{\(i \leq n\) \And \Continue}{
            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
               \lElse{\Continue\(\leftarrow\)\False}
            }
            \lOther{\Continue\(\leftarrow\)\False}

         }
         \(i\leftarrow i+1\)\;

         \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}
      }

@@

%\begin{algorithm}[]
%\begin{algorithmic}[1]
%\Function{Substitution Generation}{Input: \(\varphi\), M, P}
%   \State new\_substitutions = \(\emptyset\)
%   \ForAll {\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}
%      \ForAll {\( (\tau_1, \dots,\tau_n) \in M(f) \)}
%         \State \(\sigma\) = \{\}
%         \State continue = true
%         \While {\(i\leq n\) and continue}
%            \State subtitution\_result = match(\(\upsilon_i\), \(\tau_i\))
%            \If {substitution\_result = Success(\(\sigma_i\)) and compatible(\(\sigma_i\), \(\sigma\))}
%               \State \(\sigma = \sigma \circ \sigma_i\)
%            \Else
%               \State continue = false
%            \EndIf
%
%         \EndWhile
%
%      \EndFor
%   \EndFor
%
%   \State \Return (I, U, S)
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%In the Zipperposition implemntation this iteration procedure is limited by several bounds. The number of new type arguments that can be generated for each symbol is limited by user-defined bounds. These bounds can be defined proportionally. Additionally, the number of new type arguments can be limited at the clause level. In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.
%
%A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached. Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.
%
%Additionally, there are bounds for the number of substitutions that are generated.
%
%\begin{algorithm}[tbh]
%\begin{algorithmic}[1]
%\Function{Instantiate clause}{Clause: C, Substitutions S}
%
%   \If{\(C\) is ground}
%
%      \State \Return \(\{C\}\)
%
%   \Else
%
%      \State let \(v\) be a type variable in \(C\)
%      \State let new\_clauses be an empty set of clauses
%
%      \ForAll{subst \(\in S(C)\)}
%
%         \If{subst instantiates \(v\)}
%
%            \State \(C' = apply(subst, C)\)
%
%            \State let v\_clauses = Instantiate clause(C', S)
%
%         \EndIf
%
%         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
%
%      \EndFor
%
%      \State \Return new\_clauses
%   \EndIf
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration. In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.
%
%
%What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations. This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.
%
%\begin{algorithm}[tbh]
%\begin{algorithmic}[1]
%\Function{Generate clauses}{Problem: P, Substitutions: S}
%
%\ForAll {clause \(C \in P\)}
%   \State let new\_clauses = Instantiate clause(C, S)
%   \State \(P = P \cup new\_clauses\)
%\EndFor
%
%\State \(P' = \emptyset \)
%
%\ForAll {clause \(C \in P\)}
%   \If{\(C\) is ground}
%      \State \(P' = P' \cup \{mangle(C)\}\)
%   \EndIf
%\EndFor
%
%\State \Return \(P'\)
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%
%Implementation in Zipperposition.

>>>>

\usepackage{soul}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
\usepackage%[]
   {algorithm2e}
\usepackage{amsmath}

@@

\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{algoColorKeyword}{named}{grey}
\makeatletter
% Block with a vertical line
\renewcommand{\algocf@Vsline}[1]{%
   \strut\par\nointerlineskip%
   \algocf@bblockcode%
   \algocf@push{\skiprule}%
   \hbox{{\color{algoColorKeyword}\vrule}%
      \vtop{\algocf@push{\skiptext}%
      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
   }%
   \algocf@pop{\skiprule}%
   \algocf@bblockcode%
}
\makeatother

@@

TODO:
\begin{itemize}
   \item substitution generation
   \item clause generation
   \item fitting everything together
\end{itemize}

\begin{quote}
\begin{algorithm}[H]
   \caption{Substitution Generation}
%   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
   \Output{\,set of substitutions}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \Continue \(\leftarrow\) \True\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\) \And \Continue}{
            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
               \lElse{\Continue\(\leftarrow\)\False}
            }
            \lOther{\Continue\(\leftarrow\)\False}

            \(i\leftarrow i+1\)\;
         }

         \If{\Continue}{
            \(S\leftarrow S\cup \{\sigma\}\)\;
         }
      }

@@

Notes:
   * In the implementation, the old and new maps are actually one map with function symbols pointing to pairs of sets of type arguments
   * In the implementation, instead of tuples we have lists of types

   * In order to later have an easier time we store susbtitutions in a map that gathers them by which type variable they instantiate.

   * We assume the existence of "primitives" that compose substitutions, apply them to formulae ...

\begin{quote}
\begin{algorithm}[H]
   \caption{Clause Monomorphisation Step}

   \LinesNumbered

   \SetKwData{Subst}{substitutions}\SetKwData{TypeVars}{type\_variables}
   \SetKwData{MaxMono}{max\_mono\_args}
   \SetKwData{MaxPoly}{max\_poly\_args}
   \SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
   \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
   \SetKwData{UsedSubst}{used\_subst}

   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}

   \SetKwFunction{SubstGen}{substitution\_generation}\SetKwFunction{TyVars}{type\_variables}
   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,monomorphic type argument map \(M, M'\), new non-monomorphic type argument map \(P(\varphi), P'(\varphi)\)}
   \Output{\,monomorphic type argument map, non-monomorphic type argument map, set of substitutions}
   \BlankLine

   \Subst \(\leftarrow\) \SubstGen{\(M\), \(P(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M'\), \(P(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P'(\varphi)\)}\;
   \BlankLine

   \TypeVars \(\leftarrow\) \TyVars{\(\varphi\)}\;

   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M| \times\)\MonoMult}, \MonoCap}\;
   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)| \times\)\PolyMult}, \PolyCap}\;
   \BlankLine

   \UsedSubst \(\leftarrow\emptyset\)\;
   \ForEach{\(\sigma\in \Subst\)}{
      \ForEach{\(f\) occuring in \(\varphi\) }{
         \ForEach{\((\tau_1, \dots, \tau_n)\in P(\varphi)(f)\)}{
            \If{\TyVars{\((\tau_1, \dots, \tau_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
               \If{\(|M''| < \MaxMono\) \label{algo:relative_bound}}{
                  \(M''(f)\leftarrow M''(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }
            }
            \ElseIf{\(|P''(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
               \(P''(\varphi)(f)\leftarrow P''(\varphi)(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
            }

         }
      }
   }
   \BlankLine

   \Return \((M'', P''(\varphi), \UsedSubst)\)

\end{algorithm}
\end{quote}

At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), if we add conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), it is possible to create more fine grained bounds. This may be useful if some function symbols have many type arguments relative to the overall bound thus preventing new type arguments of other function symbols from being generated. Then, adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments.
%\begin{algorithm}[]
%\begin{algorithmic}[1]
%\Function{Substitution Generation}{Input: \(\varphi\), M, P}
%   \State new\_substitutions = \(\emptyset\)
%   \ForAll {\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}
%      \ForAll {\( (\tau_1, \dots,\tau_n) \in M(f) \)}
%         \State \(\sigma\) = \{\}
%         \State continue = true
%         \While {\(i\leq n\) and continue}
%            \State subtitution\_result = match(\(\upsilon_i\), \(\tau_i\))
%            \If {substitution\_result = Success(\(\sigma_i\)) and compatible(\(\sigma_i\), \(\sigma\))}
%               \State \(\sigma = \sigma \circ \sigma_i\)
%            \Else
%               \State continue = false
%            \EndIf
%
%         \EndWhile
%
%      \EndFor
%   \EndFor
%
%   \State \Return (I, U, S)
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%In the Zipperposition implemntation this iteration procedure is limited by several bounds. The number of new type arguments that can be generated for each symbol is limited by user-defined bounds. These bounds can be defined proportionally. Additionally, the number of new type arguments can be limited at the clause level. In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.
%
%A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached. Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.
%
%Additionally, there are bounds for the number of substitutions that are generated.
%
%\begin{algorithm}[tbh]
%\begin{algorithmic}[1]
%\Function{Instantiate clause}{Clause: C, Substitutions S}
%
%   \If{\(C\) is ground}
%
%      \State \Return \(\{C\}\)
%
%   \Else
%
%      \State let \(v\) be a type variable in \(C\)
%      \State let new\_clauses be an empty set of clauses
%
%      \ForAll{subst \(\in S(C)\)}
%
%         \If{subst instantiates \(v\)}
%
%            \State \(C' = apply(subst, C)\)
%
%            \State let v\_clauses = Instantiate clause(C', S)
%
%         \EndIf
%
%         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
%
%      \EndFor
%
%      \State \Return new\_clauses
%   \EndIf
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration. In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.
%
%
%What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations. This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.
%
%\begin{algorithm}[tbh]
%\begin{algorithmic}[1]
%\Function{Generate clauses}{Problem: P, Substitutions: S}
%
%\ForAll {clause \(C \in P\)}
%   \State let new\_clauses = Instantiate clause(C, S)
%   \State \(P = P \cup new\_clauses\)
%\EndFor
%
%\State \(P' = \emptyset \)
%
%\ForAll {clause \(C \in P\)}
%   \If{\(C\) is ground}
%      \State \(P' = P' \cup \{mangle(C)\}\)
%   \EndIf
%\EndFor
%
%\State \Return \(P'\)
%
%\EndFunction
%\end{algorithmic}
%\end{algorithm}
%
%Implementation in Zipperposition.
